---
title: "GAP analyses"
author: "Théophile L. Mouton"
date: "October 9, 2024"
format:
  html:
    toc: true
    toc-location: right
    css: custom.css
    output-file: "GAP_analyses_EEZs.html"
    self-contained: true
    code-fold: true
    code-tools: true
editor: visual
execute:
  warning: false
  message: false
  echo: true
---

# Load the data

Load the grid and the MPA raster layers

```{r}
# Load necessary libraries
library(raster)
library(here)
library(sp)
library(dplyr)
library(tidyr)

# Load the data ----
load(here::here("Data", "GAP analyses", "puvsp_marine.Rdata"))

# Load the raster layers
mpa_ALL <- raster::raster(here::here("Data", "GAP analyses", "mpa_ALL_binary.tif"))
mpa_NT <- raster::raster(here::here("Data", "GAP analyses", "mpa_NT_binary.tif"))

```

# GAP analyses

GAP analyses of shark and ray range overlaps with Marine Protected Areas

```{r}
# Convert species dataframe to spatial points
species_points <- SpatialPointsDataFrame(coords = puvsp_marine[, c("lon", "lat")], 
                                         data = puvsp_marine, 
                                         proj4string = CRS(projection(mpa_ALL)))

# Extract MPA values for each species point
mpa_ALL_values <- raster::extract(mpa_ALL, species_points)
mpa_NT_values <- raster::extract(mpa_NT, species_points)

# Add MPA values to the species dataframe
puvsp_marine$mpa_ALL_present <- mpa_ALL_values
puvsp_marine$mpa_NT_present <- mpa_NT_values

# Function to calculate percentage of range in MPA
calculate_mpa_percentage <- function(species_column, mpa_column) {
  species_presence <- species_column == 1  # Assuming 1 indicates species presence
  total_range <- sum(species_presence, na.rm = TRUE)
  range_in_mpa <- sum(species_presence & mpa_column == 1, na.rm = TRUE)
  
  if (total_range == 0) {
    return(NA)  # Return NA if the species is not present anywhere
  }
  
  percentage <- (range_in_mpa / total_range) * 100
  return(percentage)
}

# Apply function to all species columns for both MPA types
species_columns <- 4:(ncol(puvsp_marine) - 2)  # Assuming species columns start at 4
mpa_ALL_percentages <- sapply(puvsp_marine[, species_columns], 
                              function(x) calculate_mpa_percentage(x, puvsp_marine$mpa_ALL_present))
mpa_NT_percentages <- sapply(puvsp_marine[, species_columns], 
                             function(x) calculate_mpa_percentage(x, puvsp_marine$mpa_NT_present))

# Create a dataframe with results
results <- data.frame(
  species = names(mpa_ALL_percentages),
  percentage_in_ALL_MPA = mpa_ALL_percentages,
  percentage_in_NT_MPA = mpa_NT_percentages
)

# Remove any NA results
results <- results[!is.na(results$percentage_in_ALL_MPA) & !is.na(results$percentage_in_NT_MPA), ]

# Sort results by percentage in ALL MPAs (descending)
results <- results[order(-results$percentage_in_ALL_MPA), ]

# Display top 10 species
#print(head(results, 10))

# Summary statistics
cat("\nSummary Statistics for ALL MPAs:\n")
print(summary(results$percentage_in_ALL_MPA))
print(sd(results$percentage_in_ALL_MPA))

cat("\nSummary Statistics for No-Take MPAs:\n")
print(summary(results$percentage_in_NT_MPA))
print(sd(results$percentage_in_NT_MPA))

# Save results to CSV
#write.csv(results, "species_mpa_coverage_ALL_and_NT.csv", row.names = FALSE)
```

# Null model of MPA placement within EEZs

Null model of MPA placement and overlaps with the range of sharks and rays

```{r}
library(raster)
library(here)
library(sp)
library(sf)
library(parallel)
library(pbapply)

# Load the bathymetry raster
Bathy <- raster(here::here("Data", "bathymetry-0.1deg-adjusted.tif"))

# Create marine mask from bathymetry (areas < 0)
marine_mask <- Bathy < 0

# Resample marine mask to match MPA raster resolution
marine_mask_resampled <- raster::resample(marine_mask, mpa_ALL, method = "ngb")

# Ensure marine_mask_resampled is binary (0 or 1)
marine_mask_resampled <- raster::reclassify(marine_mask_resampled, c(-Inf, 0.5, 0, 0.5, Inf, 1))

# Load the EEZ shapefile
eez <- st_read(here("Data", "World_EEZ_v12_20231025", "eez_v12.shp"), quiet = TRUE)

# Function to create a raster mask of EEZs with unique values for each country
create_eez_mask <- function(template_raster) {
  # Ensure EEZ has a unique identifier for each country
  eez$country_id <- as.numeric(as.factor(eez$SOVEREIGN1))
  
  # Rasterize the EEZ shapefile using the country_id
  eez_raster <- rasterize(eez, template_raster, field = "country_id")
  
  return(eez_raster)
}

# Updated create_random_mpa function to randomize within each country's EEZ
create_random_mpa <- function(template_raster, marine_mask, eez_mask) {
  # Ensure template_raster is binary
  template_raster <- raster::reclassify(template_raster, c(-Inf, 0.5, 0, 0.5, Inf, 1))
  
  # Combine marine_mask and eez_mask
  combined_mask <- marine_mask * (eez_mask > 0)
  
  # Get unique country IDs
  country_ids <- unique(eez_mask[!is.na(eez_mask)])
  country_ids <- country_ids[country_ids > 0]
  
  # Initialize random raster
  random_raster <- template_raster
  random_raster[] <- NA
  
  for (country_id in country_ids) {
    # Create mask for current country
    country_mask <- eez_mask == country_id
    
    # Count original marine MPA cells within current country's EEZ
    original_mpa_cells <- sum(template_raster[] == 1 & country_mask[] & !is.na(combined_mask[]), na.rm = TRUE)
    
    # Get all valid cells for current country (marine areas within EEZ)
    valid_cells <- which(country_mask[] & !is.na(combined_mask[]))
    total_valid_cells <- length(valid_cells)
    
    if (original_mpa_cells > 0 && total_valid_cells > 0) {
      if (original_mpa_cells > total_valid_cells) {
        warning(paste("More MPA cells than valid marine cells for country ID", country_id, ". Adjusting MPA cell count."))
        original_mpa_cells <- total_valid_cells
      }
      
      # Randomly select valid cells to be MPAs for current country
      mpa_cells <- sample(valid_cells, size = min(original_mpa_cells, total_valid_cells), replace = FALSE)
      
      # Update random raster
      random_raster[valid_cells] <- 0
      random_raster[mpa_cells] <- 1
    }
  }
  
  return(random_raster)
}

# Calculate MPA percentage function
calculate_mpa_percentage <- function(species_values, mpa_values) {
  total_cells <- sum(!is.na(species_values))
  mpa_cells <- sum(mpa_values == 1 & !is.na(species_values), na.rm = TRUE)
  percentage <- (mpa_cells / total_cells) * 100
  return(percentage)
}

# Main analysis function
run_random_mpa_analysis <- function(species_data, mpa_raster, marine_mask, eez_mask, n_iterations = 100) {
  # Convert species dataframe to spatial points
  species_points <- sp::SpatialPointsDataFrame(coords = species_data[, c("lon", "lat")], 
                                               data = species_data, 
                                               proj4string = sp::CRS(raster::projection(mpa_raster)))
  
  # Prepare results storage
  species_columns <- 4:(ncol(species_data) - 1)  # Adjust if needed
  all_results <- matrix(nrow = length(species_columns), ncol = n_iterations)
  rownames(all_results) <- names(species_data)[species_columns]
  
  # Use pblapply for parallel processing with a progress bar
  all_results <- pblapply(1:n_iterations, function(i) {
    set.seed(i)  # Set seed for reproducibility
    random_mpa <- create_random_mpa(mpa_raster, marine_mask, eez_mask)
    random_mpa_values <- raster::extract(random_mpa, species_points)
    sapply(species_data[, species_columns], 
           function(x) calculate_mpa_percentage(x, random_mpa_values))
  }, cl = detectCores() - 1)  # Use one less than available cores
  
  all_results <- do.call(cbind, all_results)
  
  # Calculate mean and standard deviation for each species
  mean_percentages <- rowMeans(all_results, na.rm = TRUE)
  sd_percentages <- apply(all_results, 1, sd, na.rm = TRUE)
  
  results <- data.frame(
    species = rownames(all_results),
    mean_percentage = mean_percentages,
    sd_percentage = sd_percentages
  )
  
  return(results)
}

# Function to process results
process_results <- function(results, mpa_type) {
  # Sort results by mean_percentage in descending order
  results <- results[order(-results$mean_percentage), ]
  
  # Calculate summary statistics
  summary_stats <- summary(results$mean_percentage)
  
  # Write results to CSV
  write.csv(results, here::here("Outputs",paste0("species_random_", mpa_type, "_coverage.csv")), row.names = FALSE)
  
  # Return a list containing the processed data
  return(list(
    top_10 = head(results, 10),
    summary_stats = summary_stats,
    all_results = results
  ))
}

# Create the EEZ mask with unique country IDs
eez_mask <- create_eez_mask(mpa_ALL)

# Run the analysis with the country-specific EEZ constraint
tryCatch({
  random_results_ALL <- run_random_mpa_analysis(puvsp_marine, mpa_ALL, marine_mask_resampled, eez_mask, n_iterations = 100)
  print("Analysis for ALL MPAs completed successfully.")
}, error = function(e) {
  print(paste("Error in ALL MPAs analysis:", e$message))
})

tryCatch({
  random_results_NT <- run_random_mpa_analysis(puvsp_marine, mpa_NT, marine_mask_resampled, eez_mask, n_iterations = 100)
  print("Analysis for No-Take MPAs completed successfully.")
}, error = function(e) {
  print(paste("Error in No-Take MPAs analysis:", e$message))
})

# Process results for both MPA types
processed_results <- list()

if (exists("random_results_ALL")) {
  processed_results[["ALL MPAs"]] <- process_results(random_results_ALL, "ALL MPAs")
}

if (exists("random_results_NT")) {
  processed_results[["No-take MPAs"]] <- process_results(random_results_NT, "No-take MPAs")
}
```

**Important note:** This NULL model randomly distributes protected grid cells at a 0.5-degree resolution within country's EEZs. However, it does not preserve MPA shape.

# Compare results

Compare results between the MPA network and the Null model of MPA placement

```{r}
# Compare results ---- 
library(dplyr)
library(tidyr)
library(ggplot2)

# Load actual MPA coverage data
actual_coverage <- read.csv(here::here("Outputs", "species_mpa_coverage_ALL_and_NT.csv"))
colnames(actual_coverage)[2:3]=c("ALL","NT")
# Load random MPA coverage results
random_ALL <- read.csv(here::here("Outputs", "species_random_ALL MPAs_coverage.csv"))
random_NT <- read.csv(here::here("Outputs", "species_random_No-take MPAs_coverage.csv"))

# Reshape actual coverage data to long format
actual_long <- actual_coverage %>%
  pivot_longer(cols = c(ALL, NT), 
               names_to = "mpa_type", 
               values_to = "actual_percentage")

# Function to merge and compare actual vs random data
compare_coverage <- function(actual_data, random_data, mpa_type) {
  comparison <- actual_data %>%
    filter(mpa_type == !!mpa_type) %>%
    left_join(random_data, by = "species") %>%
    mutate(difference = actual_percentage - mean_percentage,
           z_score = (actual_percentage - mean_percentage) / sd_percentage)
  return(comparison)
}

# Create comparison dataframes
comparison_ALL <- compare_coverage(actual_long, random_ALL, "ALL")
comparison_NT <- compare_coverage(actual_long, random_NT, "NT")

# Function to summarize and print results
library(knitr)
library(kableExtra)

summarize_results <- function(comparison_data, mpa_type) {
  over_represented <- mean(comparison_data$difference > 0, na.rm=TRUE) * 100
  under_represented <- mean(comparison_data$difference < 0, na.rm=TRUE) * 100
  equally_represented <- 100 - over_represented - under_represented
  
  summary_df <- data.frame(
    Representation = c("Over-represented", "Under-represented", "Equally represented"),
    Percentage = round(c(over_represented, under_represented, equally_represented), 2)
  )
  
  kable(summary_df, format = "html", col.names = c("Representation", "Percentage (%)"),
        caption = paste("Summary for", mpa_type, "MPAs")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  full_width = FALSE) %>%
    column_spec(2, width = "100px") %>%
    row_spec(0, bold = TRUE) %>%
    add_header_above(c(" " = 1, "Species" = 1)) %>%
    footnote(general = "Percentages may not sum to 100% due to rounding.")
}

# Summarize results
summarize_results(comparison_ALL, "ALL")
summarize_results(comparison_NT, "No-Take")

# Identify significantly over/under-represented species
library(dplyr)

significant_species <- function(comparison_data, mpa_type, z_threshold = 1.96) {
  over_represented <- comparison_data %>% 
    filter(z_score > z_threshold) %>% 
    arrange(desc(z_score))
  
  under_represented <- comparison_data %>% 
    filter(z_score < -z_threshold) %>% 
    arrange(z_score)
  
  top_5_over <- over_represented %>% 
    select(species, actual_percentage, mean_percentage, difference, z_score) %>% 
    head(5)
  
  top_5_under <- under_represented %>% 
    select(species, actual_percentage, mean_percentage, difference, z_score) %>% 
    head(5)
  
  combined_table <- bind_rows(
    mutate(top_5_over, representation = "Over-represented"),
    mutate(top_5_under, representation = "Under-represented")
  ) %>%
    mutate(across(where(is.numeric), ~round(., 2)))
  
  kable_output <- kable(combined_table, format = "html", 
        col.names = c("Species", "Actual %", "Random %", "Difference", "Z-score", "Representation"),
        caption = paste("Top 5 Significantly Over/Under-represented Species in", mpa_type, "MPAs")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  full_width = FALSE) %>%
    column_spec(1, width = "200px") %>%
    column_spec(2:5, width = "100px") %>%
    column_spec(6, width = "150px") %>%
    row_spec(0, bold = TRUE) %>%
    pack_rows("Over-represented", 1, 5, label_row_css = "background-color: #e6f3ff; color: #000;") %>%
    pack_rows("Under-represented", 6, 10, label_row_css = "background-color: #fff0e6; color: #000;")
  
  footnote_text <- paste(
    "Total significantly over-represented species:", nrow(over_represented), "\n",
    "Total significantly under-represented species:", nrow(under_represented)
  )
  
  kable_output %>%
    footnote(general = footnote_text, general_title = "Note:")
}

# Identify significant species
significant_species(comparison_ALL, "ALL")
significant_species(comparison_NT, "No-Take")
```

**Key message:** \>60% of species are under-represented by the global MPA network (i.e. less protected by the current MPA network than by a random placement of MPAs).

# Map results of the GAP analyses

Map Standardised Effect Sizes of MPA coverage

```{r}
#For ALL MPAs 
difference_sp=comparison_ALL[,c(1,3,6)]

#Plot the difference 
# Load necessary libraries
library(dplyr)
library(tidyr)
library(ggplot2)

#SES:
# Step 1: Calculate SES for each species in comparison_ALL
comparison_ALL <- comparison_ALL %>%
  mutate(SES = (actual_percentage - mean_percentage) / sd_percentage)

# Assuming you have a similar species-to-grid-cell mapping like puvsp_marine
# Reshape puvsp_marine from wide to long format (if needed)
puvsp_long <- puvsp_marine %>%
  pivot_longer(cols = -c(id, lon, lat), names_to = "species", values_to = "presence") %>%
  filter(!is.na(presence) & presence == 1)  # Only keep rows where species is present

# Join SES data (comparison_ALL) to puvsp_long
combined_data <- puvsp_long %>%
  left_join(comparison_ALL, by = c("species" = "species"))

# Step 3: Group by grid cell (id, lon, lat) and calculate mean SES
mean_ses_per_cell <- combined_data %>%
  group_by(id, lon, lat) %>%
  summarise(mean_SES = mean(SES, na.rm = TRUE)) %>%
  ungroup()

# Load necessary libraries
library(ggplot2)
library(rnaturalearth)
library(sf)

# Step 1: Get land polygons from rnaturalearth
land <- ne_countries(scale = "medium", returnclass = "sf")

# Step 2: Plot the mean SES with a diverging color scale
g_ses_1 = ggplot() +
  geom_tile(data = mean_ses_per_cell, aes(x = lon, y = lat, fill = mean_SES)) +
  geom_sf(data = land, fill = "darkgrey", color = NA) +  # Add land in dark grey
  scale_fill_gradient2(
    low = "#2166ac", mid = "#f0f0f0", high = "#b2182b", midpoint = 0  # Blue-White-Red gradient, with midpoint at 0
  ) +
  coord_sf(xlim = c(-180, 180), ylim = c(-90, 90), expand = FALSE) +  # Set global coordinates
  theme_minimal() +
  theme(
    legend.position = "bottom",            # Position the legend at the bottom
    legend.title = element_text(hjust = 0.5),  # Center the legend title
    legend.key.width = unit(3, "cm"),      # Adjust width of the legend bar
    legend.key.height = unit(0.5, "cm")    # Adjust height of the legend bar
  ) +
  guides(
    fill = guide_colorbar(
      direction = "horizontal",            # Make the legend horizontal
      title.position = "top",              # Move the title to the top of the legend
      title.hjust = 0.5                    # Center the title horizontally
    )
  ) +
  labs(title = "Standardized Effect Size (SES) of Observed vs. Null MPA Coverage",
       x = "Longitude",
       y = "Latitude",
       fill = "Mean SES")
print(g_ses_1)

#For No-take MPAs 
difference_sp=comparison_NT[,c(1,3,6)]

#SES
# Step 1: Calculate SES for each species in comparison_ALL
comparison_NT <- comparison_NT %>%
  mutate(SES = (actual_percentage - mean_percentage) / sd_percentage)

# Assuming you have a similar species-to-grid-cell mapping like puvsp_marine
# Reshape puvsp_marine from wide to long format (if needed)
puvsp_long <- puvsp_marine %>%
  pivot_longer(cols = -c(id, lon, lat), names_to = "species", values_to = "presence") %>%
  filter(!is.na(presence) & presence == 1)  # Only keep rows where species is present

# Join SES data (comparison_ALL) to puvsp_long
combined_data <- puvsp_long %>%
  left_join(comparison_NT, by = c("species" = "species"))

# Step 3: Group by grid cell (id, lon, lat) and calculate mean SES
mean_ses_per_cell <- combined_data %>%
  group_by(id, lon, lat) %>%
  summarise(mean_SES = mean(SES, na.rm = TRUE)) %>%
  ungroup()

# Load necessary libraries
library(ggplot2)
library(rnaturalearth)
library(sf)

# Step 1: Get land polygons from rnaturalearth
land <- ne_countries(scale = "medium", returnclass = "sf")

# Step 2: Plot the mean SES with a diverging color scale
g_ses_2 = ggplot() +
  geom_tile(data = mean_ses_per_cell, aes(x = lon, y = lat, fill = mean_SES)) +
  geom_sf(data = land, fill = "darkgrey", color = NA) +  # Add land in dark grey
  scale_fill_gradient2(
    low = "#2166ac", mid = "#f0f0f0", high = "#b2182b", midpoint = 0  # Blue-White-Red gradient, with midpoint at 0
  ) +
  coord_sf(xlim = c(-180, 180), ylim = c(-90, 90), expand = FALSE) +  # Set global coordinates
  theme_minimal() +
  theme(
    legend.position = "bottom",            # Position the legend at the bottom
    legend.title = element_text(hjust = 0.5),  # Center the legend title
    legend.key.width = unit(3, "cm"),      # Adjust width of the legend bar
    legend.key.height = unit(0.5, "cm")    # Adjust height of the legend bar
  ) +
  guides(
    fill = guide_colorbar(
      direction = "horizontal",            # Make the legend horizontal
      title.position = "top",              # Move the title to the top of the legend
      title.hjust = 0.5                    # Center the title horizontally
    )
  ) +
  labs(title = "Standardized Effect Size (SES) of Observed vs. Null no-take MPA Coverage",
       x = "Longitude",
       y = "Latitude",
       fill = "Mean SES")
print(g_ses_2)

# Load the patchwork library for combining plots
library(patchwork)

# Combine the plots side by side
combined_plots <- g_ses_1 + g_ses_2

# Save the combined plot
# Adjust width and height as needed for your desired output
ggsave(here::here("SES_maps_comparison.png"), 
       combined_plots,
       width = 16,  # Width in inches
       height = 8,  # Height in inches
       dpi = 300)   # Resolution

```

**Key message:** Northern Pacific, Northern Atlantic, Southern Pacific, Southern Atlantic and global coastal/continental sharks and rays are less protected by MPAs than expected under a random placement of MPAs. This is even more the case for no-take MPAs. 

# Relate to IUCN status

1.  Relate the percentage of species range overlapped by MPAs to the IUCN Red List status with difference tests between categories

```{r}
IUCN <- readRDS(here::here("Data","GAP analyses","sharks_iucn_final.RDS"))
# Remove underscores from species names
IUCN$species <- gsub("_", " ", IUCN$Species)

results_IUCN=left_join(IUCN, results, by=c("species"))

library(ggplot2)
library(dplyr)

# Convert iucn_GE to a factor for better plotting
results_IUCN$iucn_GE <- as.factor(results_IUCN$iucn_GE)

results_IUCN <- results_IUCN %>%
  mutate(IUCN_status = case_when(
    iucn_GE == 0 ~ "LC",
    iucn_GE == 1 ~ "NT",
    iucn_GE == 2 ~ "VU",
    iucn_GE == 3 ~ "EN",
    iucn_GE == 4 ~ "CR",
    TRUE ~ "Unknown"
  ))

# Filter out the "Unknown" category
results_IUCN_filtered <- results_IUCN %>%
  filter(IUCN_status != "Unknown")

# Define IUCN colors and order
iucn_colors <- c(
  "CR" = "#D81E05",  # Red
  "EN" = "#FC7F3F",  # Orange
  "VU" = "#FEC748",  # Yellow
  "NT" = "#58AFFF",  # Light Blue
  "LC" = "#38AB38"   # Green
)

iucn_order <- c("LC", "NT", "VU", "EN", "CR")

library(ggplot2)
library(dplyr)
library(dunn.test)

# Perform Kruskal-Wallis test
kruskal_result <- kruskal.test(percentage_in_ALL_MPA ~ IUCN_status, data = results_IUCN_filtered)

# Perform Dunn's test for pairwise comparisons and capture the output
invisible(capture.output(
  dunn_result <- dunn.test(results_IUCN_filtered$percentage_in_ALL_MPA, 
                           results_IUCN_filtered$IUCN_status, 
                           method = "bonferroni")
))

# Create a function to add significance levels to the plot
add_significance <- function(p.value) {
  if(p.value <= 0.001) return("***")
  else if(p.value <= 0.01) return("**")
  else if(p.value <= 0.05) return("*")
  else return("ns")
}

# Create a data frame with pairwise comparisons and their significance
significant_comparisons <- data.frame(
  group1 = sapply(strsplit(dunn_result$comparisons, " - "), `[`, 1),
  group2 = sapply(strsplit(dunn_result$comparisons, " - "), `[`, 2),
  p.value = dunn_result$P.adjusted,
  stringsAsFactors = FALSE
)

# Add significance levels to the data frame
significant_comparisons$significance <- sapply(significant_comparisons$p.value, add_significance)

# Filter out non-significant comparisons
significant_comparisons <- significant_comparisons[significant_comparisons$significance != "ns", ]

# Calculate x positions for the significance bars
iucn_levels <- c("LC", "NT", "VU", "EN", "CR")
x_positions <- seq_along(iucn_levels)
names(x_positions) <- iucn_levels

significant_comparisons$x1 <- x_positions[significant_comparisons$group1]
significant_comparisons$x2 <- x_positions[significant_comparisons$group2]

# Sort comparisons based on the difference between group indices
significant_comparisons$group_diff <- abs(significant_comparisons$x2 - significant_comparisons$x1)
significant_comparisons <- significant_comparisons[order(significant_comparisons$group_diff, decreasing = TRUE), ]

# Calculate y positions for the sorted comparisons
significant_comparisons$y.position <- seq(max(results_IUCN_filtered$percentage_in_ALL_MPA, na.rm = TRUE) + 5, 
                                          by = 5, 
                                          length.out = nrow(significant_comparisons))

# Recreate the base violin plot
violin_plot <- ggplot(results_IUCN_filtered, aes(x = factor(IUCN_status, levels = iucn_order), y = percentage_in_ALL_MPA)) +
  geom_violin(aes(fill = IUCN_status, color = IUCN_status), trim = FALSE, alpha = 0.5) +
  geom_jitter(width = 0.1, size = 0.5, alpha = 0.5, color = "darkgray") +
  geom_boxplot(width = 0.1, fill = "white", color = "black", outlier.shape = NA, alpha = 0.8) +
  labs(x = "IUCN Status", y = "(%) Range within MPAs") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  scale_x_discrete(limits = iucn_order) +
  scale_fill_manual(values = iucn_colors) +
  scale_color_manual(values = iucn_colors)

# Add significance bars and labels
violin_plot_with_significance <- violin_plot +
  geom_segment(data = significant_comparisons,
               aes(x = x1, xend = x2, y = y.position, yend = y.position),
               inherit.aes = FALSE,
               color = "black", size = 0.5) +
  geom_segment(data = significant_comparisons,
               aes(x = x1, xend = x1, y = y.position, yend = y.position - 2),
               inherit.aes = FALSE,
               color = "black", size = 0.5) +
  geom_segment(data = significant_comparisons,
               aes(x = x2, xend = x2, y = y.position, yend = y.position - 2),
               inherit.aes = FALSE,
               color = "black", size = 0.5) +
  geom_text(data = significant_comparisons,
            aes(x = (x1 + x2) / 2, y = y.position, label = significance),
            inherit.aes = FALSE,
            vjust = -0.5, size = 3)

# Display the updated plot
print(violin_plot_with_significance)

# Save the plot
#ggsave("violin_plot_with_significance.pdf", violin_plot_with_significance, width = 10, height = 8, units = "in")

# Summary statistics
summary_stats <- results_IUCN_filtered %>%
  group_by(IUCN_status) %>%
  summarise(
    count = n(),
    mean = mean(percentage_in_ALL_MPA, na.rm = TRUE),
    median = median(percentage_in_ALL_MPA, na.rm = TRUE),
    sd = sd(percentage_in_ALL_MPA, na.rm = TRUE)
  ) %>%
  arrange(factor(IUCN_status, levels = iucn_order))

print(summary_stats)

#Now for no-take MPAs 
kruskal_result <- kruskal.test(sqrt(percentage_in_NT_MPA) ~ IUCN_status, data = results_IUCN_filtered)

# Perform Dunn's test for pairwise comparisons
# Perform Dunn's test for pairwise comparisons and capture the output
invisible(capture.output(
  dunn_result <- dunn.test(sqrt(results_IUCN_filtered$percentage_in_NT_MPA), 
                           results_IUCN_filtered$IUCN_status, 
                           method = "bonferroni")
))

# Create a function to add significance levels to the plot
add_significance <- function(p.value) {
  if(p.value <= 0.001) return("***")
  else if(p.value <= 0.01) return("**")
  else if(p.value <= 0.05) return("*")
  else return("ns")
}

# Create a data frame with pairwise comparisons and their significance
significant_comparisons <- data.frame(
  group1 = sapply(strsplit(dunn_result$comparisons, " - "), `[`, 1),
  group2 = sapply(strsplit(dunn_result$comparisons, " - "), `[`, 2),
  p.value = dunn_result$P.adjusted,
  stringsAsFactors = FALSE
)

# Add significance levels to the data frame
significant_comparisons$significance <- sapply(significant_comparisons$p.value, add_significance)

# Filter out non-significant comparisons
significant_comparisons <- significant_comparisons[significant_comparisons$significance != "ns", ]

# Calculate x positions for the significance bars
iucn_levels <- c("LC", "NT", "VU", "EN", "CR")
x_positions <- seq_along(iucn_levels)
names(x_positions) <- iucn_levels

significant_comparisons$x1 <- x_positions[significant_comparisons$group1]
significant_comparisons$x2 <- x_positions[significant_comparisons$group2]

# Sort comparisons based on the difference between group indices
significant_comparisons$group_diff <- abs(significant_comparisons$x2 - significant_comparisons$x1)
significant_comparisons <- significant_comparisons[order(significant_comparisons$group_diff, decreasing = TRUE), ]

# Calculate y positions for the sorted comparisons
significant_comparisons$y.position <- seq(max(results_IUCN_filtered$percentage_in_ALL_MPA, na.rm = TRUE) + 5, 
                                          by = 5, 
                                          length.out = nrow(significant_comparisons))

# Recreate the base violin plot
violin_plot <- ggplot(results_IUCN_filtered, aes(x = factor(IUCN_status, levels = iucn_order), y = percentage_in_NT_MPA)) +
  geom_violin(aes(fill = IUCN_status, color = IUCN_status), trim = FALSE, alpha = 0.5) +
  geom_jitter(width = 0.1, size = 0.5, alpha = 0.5, color = "darkgray") +
  geom_boxplot(width = 0.1, fill = "white", color = "black", outlier.shape = NA, alpha = 0.8) +
  labs(x = "IUCN Status", y = "(%) Range within no-take MPAs") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none") +
  scale_x_discrete(limits = iucn_order) +
  scale_y_continuous(transform = "sqrt") +
  scale_fill_manual(values = iucn_colors) +
  scale_color_manual(values = iucn_colors)

# Add significance bars and labels
violin_plot_with_significance <- violin_plot +
  geom_segment(data = significant_comparisons,
               aes(x = x1, xend = x2, y = y.position, yend = y.position),
               inherit.aes = FALSE,
               color = "black", size = 0.5) +
  geom_segment(data = significant_comparisons,
               aes(x = x1, xend = x1, y = y.position, yend = y.position - 2),
               inherit.aes = FALSE,
               color = "black", size = 0.5) +
  geom_segment(data = significant_comparisons,
               aes(x = x2, xend = x2, y = y.position, yend = y.position - 2),
               inherit.aes = FALSE,
               color = "black", size = 0.5) +
  geom_text(data = significant_comparisons,
            aes(x = (x1 + x2) / 2, y = y.position, label = significance),
            inherit.aes = FALSE,
            vjust = -0.5, size = 3)

# Display the updated plot
print(violin_plot_with_significance)

# Summary statistics
summary_stats <- results_IUCN_filtered %>%
  group_by(IUCN_status) %>%
  summarise(
    count = n(),
    mean = mean(percentage_in_NT_MPA, na.rm = TRUE),
    median = median(percentage_in_NT_MPA, na.rm = TRUE),
    sd = sd(percentage_in_NT_MPA, na.rm = TRUE)
  ) %>%
  arrange(factor(IUCN_status, levels = iucn_order))

print(summary_stats)

```

**Key message:** Least concerned species are significantly more protected by all MPAs than any other IUCN Red List threat status category. We also note a similar relationship (LC vs VU, EN & CR) for no-take MPAs, however, after p-value adjustment for multiple comparison, differences were not significant for no-take MPAs.

# Relate to species traits

Relate the percentage of species range overlapped by MPAs to the trait space of species. 1) Use Pearson correlation tests between the axes of the trait space and MPA coverage. 2) Predict relationships from a GAM into the kernel density gridded trait space.

```{r}
# Relate to species traits ----
load(here::here("Data", "GAP analyses", "coords_1005.Rdata"))
load(here::here("Data", "GAP analyses", "grids_commonspecies_corrected_021024.Rdata"))

# Load the necessary library
library(tibble)
# Convert to tibble and add row names as the first column
coords <- tibble::rownames_to_column(coords, var = "Species")

colnames(results)[1]="Species"

results_coords=left_join(coords, results)


library(Hmisc)
library(dplyr)
library(tidyr)

# Calculate correlation matrix with p-values
cor_matrix_with_p <- rcorr(as.matrix(results_coords[, c("A1", "A2", "A3", "percentage_in_ALL_MPA", "percentage_in_NT_MPA")]))

# Extract correlation coefficients and p-values
cor_coefficients <- cor_matrix_with_p$r
p_values <- cor_matrix_with_p$P

# Create a function to format the results
format_cor_p <- function(cor, p) {
  sprintf("%.3f (p = %.3f)", cor, p)
}

# Create a data frame with formatted results
result_df <- data.frame(
  Predictor = c("A1", "A2", "A3"),
  `percentage_in_ALL_MPA` = format_cor_p(cor_coefficients["percentage_in_ALL_MPA", c("A1", "A2", "A3")], 
                                         p_values["percentage_in_ALL_MPA", c("A1", "A2", "A3")]),
  `percentage_in_NT_MPA` = format_cor_p(cor_coefficients["percentage_in_NT_MPA", c("A1", "A2", "A3")], 
                                        p_values["percentage_in_NT_MPA", c("A1", "A2", "A3")])
)

library(knitr)
library(kableExtra)

format_correlation_table <- function(result_df) {
  kable(result_df, format = "html", escape = FALSE,
        col.names = c("Predictor", "ALL MPA", "NT MPA"),
        caption = "Pearson Correlation Results") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                  full_width = FALSE) %>%
    column_spec(1, bold = TRUE) %>%
    add_header_above(c(" " = 1, "Percentage in MPA" = 2)) %>%
    footnote(general = "Values shown as: correlation coefficient (p-value)",
             general_title = "Note:")
}

# Usage:
formatted_table <- format_correlation_table(result_df)
formatted_table

library(gridExtra)

# Scatter plot for A1 vs percentage in ALL MPA
plot_A1 <- ggplot(results_coords, aes(x = A1, y = percentage_in_ALL_MPA)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  theme_minimal() +
  labs(title = "A1 vs Percentage in ALL MPA",
       x = "A1",
       y = "Percentage in ALL MPA")

# Scatter plot for A2 vs percentage in ALL MPA
plot_A2 <- ggplot(results_coords, aes(x = A2, y = percentage_in_ALL_MPA)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  theme_minimal() +
  labs(title = "A2 vs Percentage in ALL MPA",
       x = "A2",
       y = "Percentage in ALL MPA")

# Set a common aspect ratio for both plots
plot_A1 <- plot_A1 + theme(aspect.ratio = 1)
plot_A2 <- plot_A2 + theme(aspect.ratio = 1)

# Create the combined plot
combined_plot <- grid.arrange(plot_A1, plot_A2, ncol = 2)
#Create kernel density of the trait space 
# Load required packages
library(BAT)

# Assuming 'coords' is your dataframe
# Extract the first two columns (A1 and A2)
load(here::here("Data", "GAP analyses", "coords_1005.Rdata"))
trait_space <- coords[, 1:2]
sp_df=grids.fd_new

# Get the species names from the trait space
trait_species <- rownames(coords)

# Subset the sp_df to keep only the columns (species) found in the trait space
sp_df_filtered <- sp_df[, colnames(sp_df) %in% trait_species]

# Replace all NA values with 0
sp_df_filtered[is.na(sp_df_filtered)] <- 0

# Create a global community matrix
global_comm <- matrix(1, nrow = 1, ncol = ncol(sp_df_filtered))
colnames(global_comm) <- colnames(sp_df_filtered)
rownames(global_comm) <- "global"

global_kernel <- 
  BAT::kernel.build(comm = global_comm, 
                    trait = trait_space,
                    method = "gaussian")

# Extract coordinates (trait values)
trait_coords <- global_kernel@Data

# Extract random points and their corresponding density values
random_points <- global_kernel@RandomPoints
density_values <- global_kernel@ValueAtRandomPoints

# Create a data frame for the density plot
plot_data <- data.frame(
  A1 = random_points[,1],
  A2 = random_points[,2],
  Density = density_values
)

# Create a data frame for the original trait points
trait_data <- data.frame(
  A1 = trait_coords[,1],
  A2 = trait_coords[,2]
)

#Build the GAM 
library(mgcv)
library(dplyr)

# Prepare data for GAM
gam_data <- results_coords %>%
  select(Species, A1, A2, percentage_in_ALL_MPA) %>%
  filter(!is.na(percentage_in_ALL_MPA))  # Remove any rows with NA in the response variable

# Build GAM
gam_model <- gam(percentage_in_ALL_MPA ~ s(A1, A2), data = gam_data, method = "REML")
summary(gam_model)

# Make predictions
plot_data$predicted <- predict(gam_model, newdata = plot_data, type = "response")

library(ggplot2)
library(viridis)
# Plot predictions
gam_plot <- ggplot() +
  geom_point(data = plot_data, aes(x = A1, y = A2, color = predicted), alpha = 0.5) +
  #geom_point(data = gam_data, aes(x = A1, y = A2), color = "red", size = 2) +
  scale_color_viridis_c(name = "Predicted % MPA Coverage") +
  theme_minimal() +
  labs(title = "GAM Predictions: Percentage in ALL MPA",
       x = "Trait Axis 1",
       y = "Trait Axis 2")

print(gam_plot)

# Create a smoother surface plot
gam_density_plot <- ggplot(plot_data, aes(x = A1, y = A2, z = predicted)) +
  stat_summary_2d(fun = mean, bins = 50) +
  scale_fill_viridis_c(name = "Predicted % MPA Coverage") +
  theme_minimal() +
  labs(title = "GAM Predictions: Percentage in ALL MPA",
       x = "Trait Axis 1",
       y = "Trait Axis 2")

print(gam_density_plot)
```

**Key message:** There is a significant, relationship between the percentage of the range of sharks and rays covered by MPAs and axis 1 & 2 of the trait space. Species located in the bottom right corner of the trait space tend to be more protected, this relationship is highly significant, however, very week (R2 = 0.04, p \< 0.001).

# Relate to phylogeny

Relate the percentage of species range overlapped by MPAs to the phylogenetic tree of species, with tests of phylogenetic signal and plots of trees with the variable mapped onto the tree.

```{r}
# Relate to phylogeny ----

# For percentage range in ALL MPAs 
library(phytools)
library(here)
library(dplyr)
library(purrr)
library(ape)
library(stringr)

# Load the list of trees
load(here::here("Data", "GAP analyses", "list_tree_sharks_p.Rdata"))

# Function to modify species names
modify_species_name <- function(name) {
  str_replace(name, " ", "_")
}

# Modify species names in the results dataframe
results_IUCN_filtered <- results_IUCN_filtered %>%
  mutate(species_modified = modify_species_name(species))

compute_phylo_signal <- function(tree, data) {
  # Ensure the data is in the same order as the tree tips and remove NAs
  matched_data <- data %>%
    filter(species_modified %in% tree$tip.label) %>%
    filter(!is.na(percentage_in_ALL_MPA)) %>%
    arrange(match(species_modified, tree$tip.label))
  
  n_species_after_matching <- nrow(matched_data)
  
  # Prune the tree to match the available data
  pruned_tree <- keep.tip(tree, matched_data$species_modified)
  
  n_tips_pruned_tree <- length(pruned_tree$tip.label)
  
  # Compute phylogenetic signal using Pagel's lambda, capturing any output
  signal <- tryCatch({
    captured_output <- capture.output({
      result <- phylosig(pruned_tree, matched_data$percentage_in_ALL_MPA, method = "lambda", test = TRUE)
    })
    result$captured_output <- captured_output
    result
  }, error = function(e) {
    error_message <- paste("Error in phylosig:", e$message)
    return(list(lambda = NA, P = NA, error = error_message, captured_output = NA))
  })
  
  # Return all information
  return(list(
    lambda = signal$lambda,
    p_value = signal$P,
    n_species = n_species_after_matching,
    n_tips_pruned_tree = n_tips_pruned_tree,
    error = if(!is.null(signal$error)) signal$error else NA,
    captured_output = paste(signal$captured_output, collapse = "\n")
  ))
}
# Apply the function to all trees: using the first two only for the moment to save time
results <- map_dfr(list_tree_sharks_p[1:2], ~compute_phylo_signal(.x, results_IUCN_filtered), .id = "tree")

# Now, create the kable separately
library(knitr)
library(dplyr)
library(kableExtra)

results %>%
  select(1:4) %>%
  kable(caption = "Phylogenetic Signal Results for all MPAs",
        col.names = c("Tree", "Lambda", "P-value", "Number of Species"),
        digits = 3,
        align = c('l', 'c', 'c', 'c')) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# Compute summary statistics
summary_stats <- results %>%
  summarise(
    mean_lambda = mean(lambda, na.rm = TRUE),
    median_lambda = median(lambda, na.rm = TRUE),
    sd_lambda = sd(lambda, na.rm = TRUE),
    mean_p = mean(p_value, na.rm = TRUE),
    median_p = median(p_value, na.rm = TRUE),
    sd_p = sd(p_value, na.rm = TRUE),
    mean_n_species = mean(n_species, na.rm = TRUE),
    min_n_species = min(n_species, na.rm = TRUE),
    max_n_species = max(n_species, na.rm = TRUE)
  )

#print(summary_stats)

# For percentage range in no-take MPAs 
library(phytools)
library(here)
library(dplyr)
library(purrr)
library(ape)
library(stringr)

# Load the list of trees
load(here::here("Data", "GAP analyses", "list_tree_sharks_p.Rdata"))

# Function to modify species names
modify_species_name <- function(name) {
  str_replace(name, " ", "_")
}

# Modify species names in the results dataframe
results_IUCN_filtered <- results_IUCN_filtered %>%
  mutate(species_modified = modify_species_name(species))

compute_phylo_signal <- function(tree, data) {
  # Ensure the data is in the same order as the tree tips and remove NAs
  matched_data <- data %>%
    filter(species_modified %in% tree$tip.label) %>%
    filter(!is.na(percentage_in_NT_MPA)) %>%
    arrange(match(species_modified, tree$tip.label))
  
  n_species_after_matching <- nrow(matched_data)
  
  # Prune the tree to match the available data
  pruned_tree <- keep.tip(tree, matched_data$species_modified)
  
  n_tips_pruned_tree <- length(pruned_tree$tip.label)
  
  # Compute phylogenetic signal using Pagel's lambda, capturing any output
  signal <- tryCatch({
    captured_output <- capture.output({
      result <- phylosig(pruned_tree, matched_data$percentage_in_NT_MPA, method = "lambda", test = TRUE)
    })
    result$captured_output <- captured_output
    result
  }, error = function(e) {
    error_message <- paste("Error in phylosig:", e$message)
    return(list(lambda = NA, P = NA, error = error_message, captured_output = NA))
  })
  
  # Return all information
  return(list(
    lambda = signal$lambda,
    p_value = signal$P,
    n_species = n_species_after_matching,
    n_tips_pruned_tree = n_tips_pruned_tree,
    error = if(!is.null(signal$error)) signal$error else NA,
    captured_output = paste(signal$captured_output, collapse = "\n")
  ))
}

# Apply the function to all trees: use only the first two trees for the moment to save time
results <- map_dfr(list_tree_sharks_p[1:2], ~compute_phylo_signal(.x, results_IUCN_filtered), .id = "tree")

# Now, create the kable separately
library(knitr)
library(dplyr)
library(kableExtra)

results %>%
  select(1:4) %>%
  kable(caption = "Phylogenetic Signal Results for no-take MPAs",
        col.names = c("Tree", "Lambda", "P-value", "Number of Species"),
        digits = 3,
        align = c('l', 'c', 'c', 'c')) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# Compute summary statistics
summary_stats <- results %>%
  summarise(
    mean_lambda = mean(lambda, na.rm = TRUE),
    median_lambda = median(lambda, na.rm = TRUE),
    sd_lambda = sd(lambda, na.rm = TRUE),
    mean_p = mean(p_value, na.rm = TRUE),
    median_p = median(p_value, na.rm = TRUE),
    sd_p = sd(p_value, na.rm = TRUE),
    mean_n_species = mean(n_species, na.rm = TRUE),
    min_n_species = min(n_species, na.rm = TRUE),
    max_n_species = max(n_species, na.rm = TRUE)
  )

#print(summary_stats)


#Plot the tree with color gradient on terminal branches 
library(ggtree)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(tidytree)

# Function to get a color palette
get_color_palette <- function(n) {
  colorRampPalette(brewer.pal(9, "YlGnBu"))(n)  # Changed to YlGnBu
}

# All MPAs 
# Function to plot the circular tree
plot_circular_colored_tree <- function(tree, data) {
  # Ensure the data is in the same order as the tree tips and remove NAs
  matched_data <- data %>%
    filter(species_modified %in% tree$tip.label) %>%
    filter(!is.na(percentage_in_ALL_MPA)) %>%
    arrange(match(species_modified, tree$tip.label))
  
  # Print summary of percentage_in_ALL_MPA for debugging
  #print(summary(matched_data$percentage_in_ALL_MPA))
  
  # Prune the tree to match the available data
  pruned_tree <- ape::keep.tip(tree, matched_data$species_modified)
  
  # Create color palette
  n_colors <- 100
  color_palette <- get_color_palette(n_colors)
  
  # Convert tree to tidy format and add percentage data
  tree_data <- as_tibble(pruned_tree) %>%
    left_join(matched_data, by = c("label" = "species_modified"))
  
  # Plot the circular tree
  p <- ggtree(pruned_tree, layout="circular", aes(color=percentage_in_ALL_MPA), size =0.3) %<+% tree_data
  
  # Color the branches
  p <- p + 
    scale_color_gradientn(colours = color_palette, 
                          name = "% Range within all MPAs",
                          limits = c(0, 100),
                          na.value = "grey50") +
    theme(legend.position = "right")
  
  # Remove default labels and add a title
  p <- p + theme(plot.title = element_text(hjust = 0.5)) 
  
  return(p)
}

# Plot the first tree
tree_plot_ALL <- plot_circular_colored_tree(list_tree_sharks_p[[1]], results_IUCN_filtered)

# Display the plot
print(tree_plot_ALL)

# No-take MPAs 
# Function to plot the circular tree
plot_circular_colored_tree <- function(tree, data) {
  # Ensure the data is in the same order as the tree tips and remove NAs
  matched_data <- data %>%
    filter(species_modified %in% tree$tip.label) %>%
    filter(!is.na(percentage_in_NT_MPA)) %>%
    arrange(match(species_modified, tree$tip.label))
  
  # Print summary of percentage_in_ALL_MPA for debugging
  #print(summary(matched_data$percentage_in_NT_MPA))
  
  # Prune the tree to match the available data
  pruned_tree <- ape::keep.tip(tree, matched_data$species_modified)
  
  # Create color palette
  n_colors <- 100
  color_palette <- get_color_palette(n_colors)
  
  # Convert tree to tidy format and add percentage data
  tree_data <- as_tibble(pruned_tree) %>%
    left_join(matched_data, by = c("label" = "species_modified"))
  
  # Plot the circular tree
  p <- ggtree(pruned_tree, layout="circular", aes(color=percentage_in_NT_MPA), size =0.3) %<+% tree_data
  
  # Color the branches
  p <- p + 
    scale_color_gradientn(colours = color_palette, 
                          name = "% Range within no-take MPAs",
                          limits = c(0, 100),
                          na.value = "grey50") +
    theme(legend.position = "right")
  
  # Remove default labels and add a title
  p <- p + theme(plot.title = element_text(hjust = 0.5)) 
  
  return(p)
}

# Plot the first tree
tree_plot_NT <- plot_circular_colored_tree(list_tree_sharks_p[[1]], results_IUCN_filtered)

# Display the plot
print(tree_plot_NT)

#ggpubr::ggarrange(tree_plot_ALL, tree_plot_NT, common.legend = T, legend = "bottom")
```

**Key message:** There are significant phylogenetic signals with the percentage of MPA coverage for all MPAs (lambda= 0.12-0.17; \< 0.05), and for no-take MPAs, this relationship is almost significant (lambda = 0.04; 0.08 \> p \> 0.05). However, patterns are not obvious on the trees. Analyses were ran for two trees only but can be replicated to all 100 trees.

# Relate to FUSE and EDGE2

Relate the percentage of species range overlapped by MPAs to FUSE and EDGE2 
```{r}
library(jsonlite)
library(here)
library(ggplot2)
library(dplyr)
library(purrr)
library(patchwork)
library(knitr)
library(kableExtra)

# Read the JSON file
data <- fromJSON(here("Data", "shark_conservation_metrics_no_freshwater.json"))

# Function to process data and create plot
process_and_plot <- function(metric_name) {
  # Merge and process data
  combined_data <- data[[metric_name]]$info %>%
    left_join(results_coords, by = "Species") %>%
    mutate(across(where(is.list), unlist),
           across(where(is.character), as.factor),
           !!metric_name := pmin(pmax(!!sym(metric_name), 0), 1))
  
  # Create plot
  plot <- ggplot(combined_data, aes_string(x = metric_name, y = "percentage_in_ALL_MPA")) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    theme_minimal() +
    labs(title = paste(metric_name, "vs % Range Protected"),
         x = metric_name,
         y = "% Range Protected") +
    scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    theme(aspect.ratio = 1)
  
  # Calculate correlation
  cor_test <- cor.test(combined_data[[metric_name]], combined_data$percentage_in_ALL_MPA, use = "complete.obs")
  
  return(list(plot = plot, cor_test = cor_test))
}

# Process and plot for EDGE2 and FUSE
edge2_results <- process_and_plot("EDGE2") 
fuse_results <- process_and_plot("FUSE") 

# Combine plots in a grid with 2 columns
combined_plot <- edge2_results$plot + fuse_results$plot +
  plot_layout(ncol = 2) +
  plot_annotation(
    title = NULL,
    theme = theme(plot.title = element_text(hjust = 0.5, size = 16))
  )

# Print the combined plot
print(combined_plot)

# Create a data frame with correlation test results
correlation_results <- data.frame(
  Metric = c("EDGE2", "FUSE"),
  Correlation = c(edge2_results$cor_test$estimate, fuse_results$cor_test$estimate),
  P_value = c(edge2_results$cor_test$p.value, fuse_results$cor_test$p.value)
)

# Create and print the kable table
kable_table <- kable(correlation_results, 
                     col.names = c("Metric", "Correlation", "P-value"),
                     digits = c(0, 3, 4),
                     caption = "Correlation Test Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

kable_table
```

**Key message:** There is significant negative relationship between EDGE2 and FUSE with the % of range protected. Species with high FUSE or high EDGE2 tend to be less protected than those with low EDGE2 or FUSE. 

